---
title: "Homework 3 - Reading and Manipulating Data"
format: html
editor: visual
author: Evan Whitfield, Ryan Friedman, Ralph Church
---

## Task 1: Conceptual Questions

#### Question 1

**If your working directory is myfolder/homework/, what relative path would you specify to get the file located at myfolder/MyData.csv?**

../MyData.csv

#### Question 2

**What are the major benefits of using R projects?**

R projects provide a way to divide your work. Each with their own:

Working directory Workspace (environment with stored data/objects) History Folder structure & source documents

It is very easy to connect with GitHub for version control and collaboration.

#### Question 3

**What is Git and what is GitHub?**

Git is a version control system. GitHub is a hosting service that provides a home for our Git based projects.

#### Question 4

**What are the two main differences between a `tibble` and a `data.frame`?**

Tibbles print cleaner and usual return a tibble when you subset them. This makes them easier to chain.

Data frames try to print everything by default, but usually try to print out a vector, making them not print results as cleanly. Also, not easily chain-able.

#### Question 5

**Rewrite the following nested function call using baseR’s chaining operator:**

``` r
library(tidyverse)
arrange(filter(select(as_tibble(iris), starts_with("Petal"), Species), Petal.Length < 1.55), Species)
```

Answer:

``` r
as_tibble(iris) |>
  select(starts_with("Petal"), Species) |>
  filter(Petal.Length < 1.55) |>
  arrange(Species)
```

#### Question 6

**What is meant by long format data and wide format data? Which do we generally prefer for statistical analysis?**

Long format - Each row is a single observation Wide format - Each row is a subject, different columns are different observations

For statistical analysis, we perfer long form as it is much easier to work with for modeling and is more flexible. It is also easier to group and summarize.

## Task 2: Reading Delimited Data

We will start with Glass Data. When the data were viewed, it is shown to be delimited by a comma.

#### Glass Data

First we need to read in the tidyverse package to be prepared to read in our data as nice tibbles.

```{r}
#| message: false
#| warning: false
library(tidyverse)
```

Now, let's read in our comma separated data and assign column names in the process

```{r}
glass_data <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/glass.data",
                       col_names = c("ID",
                                     "RI", 
                                     "Na", 
                                     "Mg",
                                     "Al",
                                     "Si",
                                     "K",
                                     "Ca",
                                     "Ba",
                                     "Fe",
                                     "Type_of_Glass")
)
glass_data
```

```{r}
glass_data |> 
  mutate(
    Type_of_Glass = factor(
      recode(
        Type_of_Glass,
        '1' = "building_windows_float_processed",
        '2' = "building_windows_not_float_processed",
        '3' = "vehicle_windows_float_processed",
        '4' = "vehicle_windows_not_float_processed",
        '5' = "containers",
        '6' = "tableware",
        '7' = "headlamps"
      )
    )
  ) |>
  filter(Fe < 0.2, Type_of_Glass %in% c("tableware","headlamps"))
```

#### Yeast Data

```{r}
#| results: hide
#| message: false
yeast_data <- read_table(
  "https://www4.stat.ncsu.edu/online/datasets/yeast.data",
  col_names = c("seq_name", 
                "mcg", 
                "gvh", 
                "alm", 
                "mit", 
                "erl", 
                "pox",
                "vac", 
                "nuc", 
                "class")
)
```

```{r}
yeast_data
```

```{r}
yeast_data |> 
  select(-"seq_name", -"nuc") |>
  group_by(class) |>
  mutate(across(where(is.numeric), list(mean = mean, median = median), .names = "{.col}_{.fn}"))
```

## Task 3: Combining Excel and Delimited Data

## Task 4: Database Practice
