---
title: "Homework 3 - Reading and Manipulating Data"
format: html
editor: visual
author: Evan Whitfield, Ryan Friedman, Ralph Church
---

## Task 1: Conceptual Questions

##### Exam Practice Questions

1. If your working directory is `myfolder/homework/`, what relative path would you specify to access the file located at `myfolder/MyData.csv`?

   **Answer:** `../MyData.csv`

   Explanation:
   - `..` moves up one directory level from `homework/` to `myfolder/`.
   - `MyData.csv` is located directly inside `myfolder/`.
   
2. What are the major benefits of using R Projects?

   **Answer:**
   - Automatically sets the working directory.
   - Keeps project files organized.
   - Makes relative paths easier to use.
   - Improves reproducibility and collaboration.

3. What is Git and what is GitHub?

   **Answer:**
   - **Git** is a version control system used to track changes to files and code.
   - **GitHub** is a cloud platform that hosts Git repositories and enables collaboration.

4. What are the two main differences between a tibble and a data.frame?

   **Answer:** 
   - Tibbles print only a preview of the data instead of the entire object. 
   - Tibbles are more strict ("lazy and surly"), doing less automatic conversion and forcing you to address problems     earlier.

5. Rewrite the following nested function call using baseR’s chaining operator:

```{r}
library(tidyverse)
arrange(filter(select(as_tibble(iris), starts_with("Petal"), Species), Petal.Length <
1.55), Species)

as_tibble(iris) |>
  select(starts_with("Petal"), Species) |>
  filter(Petal.Length < 1.55) |>
  arrange(Species)
```


6. What is meant by long format data and wide format data? Which do we generally prefer for statistical analysis?

   **Answer:**

   - **Long format:** Each row consists of one observation and each column one variable.
   - **Wide format:** More than one observation may exist in a given row.

   Wide format data is often used in pivot tables and can be easier to display. However, long format data is generally preferred for statistical analysis because most plotting, summarizing, and modeling tools expect data to be in long form.

## Task 2: Reading Delimited Data

##### First, we read in the data using the URL and the **read_delim()** funcion:
```{r}
glass_data <- read_delim("https://www4.stat.ncsu.edu/~online/datasets/glass.data", 
                       delim = ",",
                       col_names = c("Id", "RI", "Na", "Mg", "Al", "Si", "K", "Ca", "Ba", "Fe", "Type_of_glass")
)

print(glass_data)
```

##### The recode() function replaces the numeric glass type codes with descriptive labels. The labels are then converted to a factor (categorical data) using factor(), and mutate() is used to save the updated values back to the Type_of_glass column.

```{r}
glass_data <- glass_data |>
  mutate(
    Type_of_glass = factor(
      recode(
        Type_of_glass,
        `1` = "building_windows_float_processed",
        `2` = "building_windows_non_float_processed",
        `3` = "vehicle_windows_float_processed",
        `4` = "vehicle_windows_non_float_processed",
        `5` = "containers",
        `6` = "tableware",
        `7` = "headlamps"
      )
    )
  ) |>
  filter(
  Fe < 0.2 &
  (Type_of_glass == "tableware" |
   Type_of_glass == "headlamps")
)

class(glass_data$Type_of_glass)

print(glass_data)
  
```

##### Yeast data:

```{r}
yeast_data <- read_delim("https://www4.stat.ncsu.edu/~online/datasets/yeast.data", 
                       delim = "  ",
                       col_names = c("seq_name", "mcg", "gvh", "alm", "mit", "erl", "pox", "vac", "nuc", "class"))
                       
yeast_data <- yeast_data |>
  select(-c(seq_name, nuc)) |>
  group_by(class) |>
  mutate(across(where(is.numeric), list(mean = mean, median = median), .names = "{.col}_{.fn}"))

print(yeast_data)
```


## Task 3: Combining Excel and Delimited Data

```{r}
library(readxl)
white_wine_data <- read_excel("white-wine.xlsx")
white_wine_data
```

##### Next, we look at the excel sheets available by calling the method **excel_sheets()**, and then subsequently use those variable names in place of the current column names using the **colnames()** method.

```{r}
excel_sheets("white-wine.xlsx")

variable_names <- read_excel("white-wine.xlsx", sheet = excel_sheets("white-wine.xlsx")[2])
variable_names
```

##### An important observation is the use of **str()** in order to identify the keys or columns that belong to the tibble, variable_names, which in this case is *Variables,* which we then use to extract the vector of variables.
```{r}
str(variable_names)

colnames(white_wine_data) <- variable_names$Variables
white_wine_data
```

##### Assign each of the rows a corresponding type of *white* indicating that they are of type white whine.

```{r}
white_wine_data$type <- "white"
```


##### Read in the red wine data and display the tibble. Next, bind the two data frames (tibbles) into one data frame using **bind_rows().**

##### Then, we want to look at taking more than one column from the dataset and the **select()** method works well when dealing with more than one column at a time, as opposed to the **dplyr::pull()** method which gets a single column.

```{r}
library(readr)
red_wine_data <- read_delim(delim = ';', "red-wine.csv")
red_wine_data

#Overwrite the current column names:
colnames(red_wine_data) <- variable_names$Variables
red_wine_data$type <- "red"

wine_data <- dplyr::bind_rows(white_wine_data, red_wine_data)
wine_data <- wine_data |>
  filter(quality > 6.5 & alcohol < 132) |>
  arrange(desc(quality)) |>
  select(citric_acid, alcohol, type, quality) |>
  group_by(quality) |>
  mutate(across(
    c(alcohol),
    list(mean = mean, sd = sd),
    .names = "{.col}_{.fn}"
  ))

wine_data
```

## Task 4: Database Practice

##### First, we connect to the Lahman SQLite database and inspect the available tables. We then demonstrate querying the database both through the tbl() interface and by writing raw SQL statements directly.

```{r}
library(Lahman)
library(dplyr)
library(DBI)

con <- dbConnect(RSQLite::SQLite(), "lahman.db")

dbListTables(con)

tbl(con, "Teams")

tbl(con, sql(
"SELECT *
FROM `Teams`
"))

tbl(
  con,
  sql("
    SELECT `nameLast`
    FROM `People`
    WHERE (playerID = 'cobbty01')
  ")
)
```
##### Next, we query the HallOfFame table and return only those players who were inducted. The results are then sorted by player ID and restricted to the columns of interest.
```{r}
tbl(con, sql(
"SELECT *
FROM `HallOfFame`
"))|>
  arrange(playerID)

tbl(con, sql(
"SELECT `playerID`, `yearID`, `category`
FROM `HallOfFame`
WHERE (`inducted` = 'Y')")
)
```
##### We then join the HallOfFame and People tables using the shared playerID column. This allows us to add each player's first and last name to the Hall of Fame records.
```{r}
tbl(con, "HallOfFame") |>
  filter(inducted == "Y") |>
  inner_join(
    tbl(con, "People"),
    by = "playerID"
  ) |>
  select(playerID, yearID, category, nameFirst, nameLast)

```
##### Next, we join the HallOfFame and Managers tables in order to identify Hall of Fame inductees who also managed teams. We then group by manager, calculate their career totals, and compute their overall win percentage.

```{r}
tbl(con, sql(
"SELECT *
FROM `Managers`
"))

tbl(con, "HallOfFame") |>
  filter(inducted == "Y") |>
  inner_join(
    tbl(con, "Managers"),
    by = "playerID"
  ) |>
select(playerID, G, W, L) |>
group_by(playerID) |>
summarize(G_managed = sum(G, na.rm = TRUE),
Total_W = sum(W, na.rm = TRUE),
Total_L = sum(L, na.rm = TRUE)) |>
  collect() |>
  mutate(
    win_percentage = Total_W / G_managed
  ) |>
  arrange(desc(win_percentage))
```
##### Finally, we combine the Hall of Fame player information with the managerial statistics. This produces a single table containing each Hall of Fame manager's induction information, name, career managerial totals, and win percentage.
```{r}
hof_people <- tbl(con, "HallOfFame") |>
  filter(inducted == "Y") |>
  inner_join(
    tbl(con, "People"),
    by = "playerID"
  ) |>
  select(playerID, yearID, category, nameFirst, nameLast)

manager_stats <- tbl(con, "HallOfFame") |>
  filter(inducted == "Y") |>
  inner_join(
    tbl(con, "Managers"),
    by = "playerID"
  ) |>
  filter(plyrMgr == "Y") |>
  group_by(playerID) |>
  summarize(
    G_managed = sum(G, na.rm = TRUE),
    Total_W = sum(W, na.rm = TRUE),
    Total_L = sum(L, na.rm = TRUE)
  ) |>
  mutate(
  win_percentage = Total_W / G_managed
)

hof_people |>
  inner_join(
    manager_stats,
    by = "playerID"
  )
```

